Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-22 bbbq_article.tex
Process feedback

In the final set of experiments, we addressed the question whether TMHs would be more evolutionary conserved than soluble proteins regions by comparing the occurrences of SNPs in the genome coding for TMHs and soluble protein regions. 
We obtained 1,129 gene names associated with the phrase 'membrane protein'.
These genes are linked to 5,203 proteins
\richel{GvdB: Why so many? Isoforms?}, of which 2,754
are predicted to be transmembrane proteins.
We obtained 4,811 SNPs that resulted in an
amino acids substitution, of which 2,568 were located 
in predicted transmembrane proteins.
Per protein, we predicted the percentages of the gene coding for TMHs and soluble regions
and calculated the percentage of amino acids substitutions
occurring in TMHs.
Per protein, this percentage was plotted as a function of the percentage TMH in figure
\ref{fig:f_snps_found_and_expected}.
We performed a linear regression analysis on the data, with and without
membrane-associated proteins \richel{GvdB: explain clearer}, and calculated the 95\%
confidence interval.
Since the regression curve clearly deviated from the line of equality,
which would be expected if SNPs occur just as likely within transmembrane and soluble protein domains,
this suggests that variations occur more often in soluble 
than in transmembrane protein regions.

To test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution. We first calculated the expected number of SNPs
found in TMHs, which is 61,457 SNPs \richel{GvdB: I don’t completely follow this. Based on what?} and multiplied this number by 6.8\% 
of all proteins being a TMH, resulting in the null hypothesis
of 4203 SNPs occurring in TMHs. 
Note that including or excluding
the membrane associated proteins 
has no effect on this statistical test, as the chance 
of a successful trial is zero (there is zero chance
to find a SNP in TMHs if there are none) \richel{GvdB: Also too complex for biologists. Explain in bit more words}.
The actually number of SNPs found in TMHs
was 3889, which is lower than the null hypothesis. 
The chance to find, within TMHs, this amount or less SNPs 
equals $1.6201 \cdot 10^{-9}$.
Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

>>>>

In the final set of experiments, we addressed the question whether TMHs would be more evolutionary conserved than soluble proteins regions by comparing the occurrences of SNPs in the genome coding for TMHs and soluble protein regions. 
We obtained 1,129 gene names associated with the phrase 'membrane protein'.
These genes are linked to 5,203 protein isoforms, 
of which 2,754 are predicted to be transmembrane proteins.
We obtained 4,811 SNPs that resulted in an
amino acids substitution, of which 2,568 were located 
in predicted transmembrane proteins.
Per protein, we calculated two percentages: 
(1) the percentage of the gene coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without membrane-associated proteins,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs and soluble protein domains.
Our analysis shows that SNPs occur more often in soluble domains
than in TMHs.
The effect size is that per 1 SNP found in soluble protein
domains, one finds 0.925 SNPs in TMHs.

To statistically test whether SNPs occur just as likely in soluble as
in transmembrane protein domains, we used a Poisson binomial
distribution.

We first calculated the expected number of SNPs
found in TMHs, which is 61,457 SNPs \richel{GvdB: I don’t completely follow this. Based on what?} and multiplied this number by 6.8\% 
of all proteins being a TMH, resulting in the null hypothesis
of 4203 SNPs occurring in TMHs. 
Note that including or excluding
the membrane associated proteins 
has no effect on this statistical test, as the chance 
of a successful trial is zero (there is zero chance
to find a SNP in TMHs if there are none) \richel{GvdB: Also too complex for biologists. Explain in bit more words}.
The actually number of SNPs found in TMHs
was 3889, which is lower than the null hypothesis. 
The chance to find, within TMHs, this amount or less SNPs 
equals $1.6201 \cdot 10^{-9}$.
Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.
